---
title: "Week 7 Exercises"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}

```

## Exercise 1: one-way and two-way ANOVAs

Consider the penguin dataset. Focus on body mass as the response variable and the sex and species of penguins as explanatory variables. As usual, subset your dataset so that any NA entries from these columns are removed.

1. Consider the penguin species as the sole explanatory variable. Determine if one-way ANOVA or Kruskal-Wallis test should be performed.

2. Perform the appropriate one-factor test based on your results from (1). Is the difference between species statistically significant?

3. Perform the post-hoc pairwise test consistent with the test you performed in (2). Which pairwise difference is significant?

4. Now consider BOTH species and sex as explanation variables, with possible interactions. Again determine which test is appropriate.

5. Conduct the appropriate test based on your results from (4). Is the interaction significant?


## Exercise 2: multiple linear regression with categorical variables

Consider the penguins dataset again. Focus on the numerical variables (body mass, bill length, bill depth, and flipper length) as well as species. As before we'll treat body mass as the response variable and the remaining variables as explanatory.

1. Suppose we are interested in the how bill depth and species may predict body mass. Fit an additive linear model, and print the summary (use the default treatment contrast). Interpret the coefficients of the summary

2. Determine if the effect of the species variable is statistically significant or not in the above model.

3. Now add the interaction term. Determine if the interaction term is statistically significant in the new model.


## Exercises 3: fixed-effect versus random-effect variables

Suppose you want to study the dependence of primary productivity on sea surface temperature using observations collected from multiple cruises across multiple ocean basins.

What are some fixed-effect variables in this study? How about random-effect variables?


## Exercise 4: Nested versus crossed effects

Suppose you are investigating how phosphate concentration changes with depth across a number of stations, and you sample each stations multiple times during the year. Due to ship availability the sampling dates do not line up, but the samples are always collected at dawn, and multiple bottles are collected for each depth every time a station is visited. If you know a priori that seasonal effect are negligible, what would be a suitable random intercept only mixed effect model for the data collected from this study?


## Exercises 5: LMM estimation on rat pups

In a study of rat pup, each litter is randomly assigned to a treatment level (control, low, high), and the birth weight and sex of each pup is measured. In addition, the weight of the litter prior to birth is also measured.

```{r}
data(RatPupWeight, package="nlme")
ratpup <- as.data.frame(RatPupWeight) |> mutate(
  Litter = factor(Litter, ordered=FALSE), 
  Treatment = factor(Treatment, ordered=FALSE)
) 

glimpse(ratpup)
```

We'll treat the the birth weight (`weight`) as response variable, and take the treatment of litter, litter size, and sex of pup as fixed factor. Finally, we'll consider the litter as random factor.

1. Fit a random intercept model in which there is no interaction between the fixed factors, and obtain its summary information

2. Which fixed-effect variables are statistically significant? (Hint: recall that significance of Treatment should be judged as a whole rather than as 2 separate effects)

3. Refit a new model where the effect of sex is also affected by the random factor of litter, and print its summary

4. How does the random effect on the intercept correlates with the random effect on sex in this new model? Illustrate the situation by making a scatter plot of the random effect coefficients


## (Optional) Feedback questions

1. Are there any topic(s) from this week that still confuse you?

2. Which activities work and which doesn't? What do you want to see change and what do you want to keep?
