---
title: "Mixed effect model estimation"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(varTestnlme)
library(broom)
library(broom.mixed)
```


## Introductory example

### Loading dataset

```{r}
fish <- read_csv("data/fishdata.csv")
glimpse(fish)
```

Narrow down to species "S1"

```{r}
fish_sub <- fish |> filter(Fish_Species=="S1")
```

Perform linear mixed model. The `(1|Lake)` means that we have a random effect term associated with the Lake factor, and the term only affect the intercept.

```{r}
fish_lmm1 <- lmer(Trophic_Pos ~ Fish_Length + (1|Lake), fish_sub)
fish_lmm1
```

Obtain the summary

```{r}
fish_lmm1_sm <- summary(fish_lmm1)
fish_lmm1_sm
```

Tidy up the information provided by model using `tidy()` from `broom.mixed`

```{r}
tidy(fish_lmm1)
```

## Including species as a fixed factor

```{r}
fish_lmm2 <- lmer(Trophic_Pos ~ Fish_Length + Fish_Species + (1|Lake), fish)
fish_lmm2_sm <- summary(fish_lmm2)
fish_lmm2_sm
```

As before we should test the Species factor as a whole, which can be done using `anova()`

```{r}
anova(fish_lmm2)
```

## Including species as a second random factor, independent of lake

```{r}
fish_lmm3 <- lmer(
  Trophic_Pos ~ Fish_Length + (1|Fish_Species) + (1|Lake), 
  fish
)
fish_lmm3_sm <- summary(fish_lmm3)
fish_lmm3_sm
```


## Including random slope, WITHOUT covariance

```{r}
fish_lmm4 <- lmer(
  Trophic_Pos ~ Fish_Length + Fish_Species + (1 + Fish_Length || Lake), 
  fish
)
fish_lmm4_sm <- summary(fish_lmm4)
fish_lmm4_sm
```

## Including random slope, WITH covariance

```{r}
fish_lmm5 <- lmer(
  Trophic_Pos ~ Fish_Length + Fish_Species + (1 + Fish_Length | Lake), 
  fish
)
fish_lmm5_sm <- summary(fish_lmm5)
fish_lmm5_sm
```

We note that in both cases the Standard deviation in the slope parameter is very small. This suggest we should stick with the intercept-only model (more on that later)....


## Extracting information

Let's return to the simpler random-intercept only model:

```{r}
fish_lmm <- lmer(
  Trophic_Pos ~ Fish_Length + Fish_Species + (1|Lake), 
  fish
)
```

Extract fixed effect

```{r}
fixef(fish_lmm)
```

Extract random effect:

```{r}
ranef(fish_lmm)
```

Extract total effect (= random effect + fixed effect)

```{r}
coef(fish_lmm)
```


```{r}
fish_lmm_A <- lmer(
  Trophic_Pos ~ Fish_Length + (1 | Fish_Species) + (1 | Lake), 
  fish
)
summary(fish_lmm_A)
```

```{r}
fish_lmm_B <- lmer(
  Trophic_Pos ~ Fish_Length + (1 | Fish_Species), 
  fish
)
summary(fish_lmm_B)
```


```{r}
varCompTest(fish_lmm_A, fish_lmm_B)
```

```{r}
anova(fish_lmm_B, fish_lmm_A)
```

