---
title: "Linear regression with categorical variables"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(broom)
library(car)
library(e1071)
library(dunn.test)
library(ARTool)
```

## One-way ANOVA

### Loading data

We'll look at the ToothGrowth dataset, an investigation of the length of odontoblasts in 60 guinea pigs. Each animal received one of three dose levels of vitamin C by one of two delivery methods, orange juice (`OJ`) or ascorbic acid (`VC`)

Loading the dataset

```{r}
tooth = datasets::ToothGrowth
```

Code the dose (which can be 1 of 3 distinct values) as a factor too

```{r}
tooth <- tooth |> mutate(dose = as.factor(dose))
```

```{r}
glimpse(tooth)
```

### One-way ANOVA on dose

Longer but more methodological way: fit an `lm()` and then perform `anova()`

```{r}
tooth_lm <- lm(len ~ dose, tooth)
tooth_av <- anova(tooth_lm)
tooth_av
```

Thus the effect of dose is significant.

Alternatively, we can use `aov()` to run ANOVA in one go:

```{r}
aov(len ~ dose, tooth)
```


### Checking assumptions

Check if there's any heteroscedasticity:

```{r}
leveneTest(tooth_lm)
```

So the null hypothesis (homogenity of variance) is not rejected.

Now let's also check for non-normality

```{r}
ggplot(tooth, aes(sample=len, color=dose)) + geom_qq() + geom_qq_line()
```

```{r}
tooth |> group_by(dose) |> summarize(
  skewness = skewness(len, type=2),
  kurtosis = kurtosis(len, type=2)
)
```

The skewness of data at dose=0.5 is somewhat notable. 

### Non-parametric alternative

Given the skewness and the small size of samples, let's run Kruskal-Wallis to make sure things are consistent even if we take a non-parametric test:

```{r}
kruskal.test(tooth$len, tooth$dose)
```

So we see that our conclusion has not changed with a non-parametric test

### Post-hoc tests

For post-hoc tests, if we ignore the skewness issue we can just run Tukey's HSD (which needs to run on an object created by `aov()`)

```{r}
tooth_tukey <- TukeyHSD(aov(len ~ dose, tooth))
tooth_tukey
```

We see that all pairwise comparison is statistically significant.

Alternatively, the non-parametric analogue of Tukey's HSD is the Dunn's test

```{r}
tooth_dunn <- dunn.test(tooth$len, tooth$dose, method="holm")
```

Thus our conclusion has not changed.

## Two-way ANOVA

Again consider the ToothGrowth dataset, where `dose` is coded as factor

### Loading data

```{r}
tooth = datasets::ToothGrowth
```

Code the dose (which can be 1 of 3 distinct values) as a factor too

```{r}
tooth <- tooth |> mutate(dose = as.factor(dose))
```

```{r}
glimpse(tooth)
```

### Two-way ANOVA

Let's perform two-way ANOVA. Again we first fit the full model and then pass it to the ANOVA function (this time we use `Anova()` from `car` and specify `type=III`). Note also that we have to set the contrasts for the categorical variables to `const.sum`

```{r}
tooth_lm_int <- lm(
  len ~ dose * supp, tooth, 
  contrast=list(dose=contr.sum, supp=contr.sum)
)
tooth_av_int <- Anova(tooth_lm_int, type="III")
print(tooth_av_int)
```

### Checking assumptions

```{r}
leveneTest(tooth_lm_int)
```

The p-value is somewhat small, so we may want to proceed with caution.

Also check normality:

```{r}
ggplot(tooth, aes(sample=len, color=dose, shape=dose)) + 
  geom_qq() + geom_qq_line() + facet_grid(~supp)
```

```{r}
tooth |> group_by(dose, supp) |> summarize(
  skewness = skewness(len, type=2),
  kurtosis = kurtosis(len, type=2)
)
```

We see some issues with skewness and kurtosis for some of the subgroup. So perhaps we should go the non-parametric route

### Non-parametric tests

```{r}
tooth_art <- art(len ~ dose * supp, tooth)
tooth_art_av <- anova(tooth_art)
tooth_art_av
```

Again we see that our conclusion has not changed, but the p-values have changed slightly in various places.


## Both categorical and continuous explanatory variables

Let's go back to the subsetted CalCOFI dataset

### Loading data

Load the subsetted data:

```{r}
CalCOFI <- read_csv("data/CalCOFI_sub.csv")
```

Inspect the dataset

```{r}
glimpse(CalCOFI)
```

## Categorical variable

There are actually 3 stations within the subsetted data

```{r}
unique(CalCOFI$station)
```

To code the stations categorical, we use as.factor

```{r}
CalCOFI2 <- CalCOFI |> mutate(station = as.factor(station))
```

Now we can perform regression with station as a factor

```{r}
O_lm_cat <- lm(O_mmol_kg ~ depth_m + salinity + T_degC + station, CalCOFI2)
```

Display the summary:

```{r}
summary(O_lm_cat)
```

In term of testing the signifcance of the station variable, what we should really do is to compare the reduced model (where `station` is absent) with the full model (where `station` is present):

```{r}
O_lm_num <- lm(O_mmol_kg ~ depth_m + salinity + T_degC, CalCOFI2)
anova(O_lm_num, O_lm_cat)
```

