---
title: "Mixed effect model diagnostics"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(varTestnlme)
library(broom)
library(broom.mixed)
```


## Loading dataset

```{r}
fish <- read_csv("data/fishdata.csv")
glimpse(fish)
```

## Fitting final model

```{r}
fish_lmm <- lmer(
  Trophic_Pos ~ Fish_Length * Fish_Species + (1 | Lake),
  fish
)
summary(fish_lmm)
```

Note that our final model has no convergence or singularity issues.

## Diagnostics

The overall residual plot

```{r}
ggplot() + geom_point(aes(x=fitted(fish_lmm), y=resid(fish_lmm))) 
```

The overall Q-Q plot:

```{r}
ggplot(mapping=aes(sample=resid(fish_lmm))) + 
  geom_qq() + geom_qq_line() 
```

Plots of residuals faceted by random factor level

```{r}
ggplot(fish) + geom_point(aes(x=fitted(fish_lmm), y=resid(fish_lmm))) +
  facet_wrap(vars(Lake))
```

Q-Q plots faceted by random factor level

```{r}
ggplot(fish, aes(sample=resid(fish_lmm))) + 
  geom_qq() + geom_qq_line() + facet_wrap(vars(Lake)) 
```

We see no strong violations on homoscedasticity or normality 

Check the distribution of random factor

```{r}
fish_ranef <- ranef(fish_lmm)$Lake
```

```{r}
ggplot(mapping=aes(sample=fish_ranef[["(Intercept)"]])) + 
  geom_qq() + geom_qq_line()
```

There is some sign of heavy tails, but that's unlikely to cause any real troubles.
