---
title: "Mixed effect model predictions"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(varTestnlme)
library(broom)
library(broom.mixed)
library(ciTools)
```

## Loading dataset

```{r}
fish <- read_csv("data/fishdata.csv")
fish_sub <- fish |> filter(Fish_Species=="S1")
glimpse(fish_sub)
```

## Fitting a simple mixed model

```{r}
fish_lmm <- lmer(
  Trophic_Pos ~ Fish_Length + (1 | Lake),
  fish_sub
)
summary(fish_lmm)
```

## Prediction about the mean response

### Marginal prediction

Creating the data frame upon which our prediction will run

```{r}
fish_all_new <- data.frame(
  Fish_Length = seq(80, 430, 10)  
)
```

Making marginal prediction

```{r}
fish_marg_pred <- predict(fish_lmm, fish_all_new, re.form=NA)
fish_marg_pred <- cbind(fish_all_new, fish_marg_pred) |>
  rename(pred = fish_marg_pred)
```

Plot the observed data with the prediction

```{r}
ggplot(fish_sub) + 
  geom_point(aes(
    x=Fish_Length, y=Trophic_Pos, 
    color=Lake, shape=Lake
  )) + geom_line(
    aes(x=Fish_Length, y=pred), fish_marg_pred
  ) 
```

### Conditional prediction

Set up the data frame for lake-specific predictions:

```{r}
lakes <- distinct(fish_sub, Lake)$Lake
lengths <- seq(80, 430, 10)

fish_new <- data.frame(
  Fish_Length = rep(lengths, each=length(lakes)),
  Lake = rep(lakes, times=length(lengths))
)
```

Make Lake-specific predictions:

```{r}
fish_cond_pred <- predict(fish_lmm, fish_new)
fish_cond_pred <- cbind(fish_new, fish_cond_pred) |>
  rename(pred = fish_cond_pred)
```

Plot the results

```{r}
ggplot(
  fish_sub, 
  aes(x=Fish_Length, y=Trophic_Pos, color=Lake, shape=Lake)
) + geom_point() +
  geom_line(
    aes(y=pred),
    data=fish_cond_pred
  ) + geom_line(
    aes(y=pred, color=NULL, shape=NULL), 
    fish_marg_pred
  ) 
```

Observe that the fitted line for L4 is closer to the unconditional prediction than the data alone would suggest. This is an example of shrinkage: our a priori knowledge (or assumption) that the lakes factor are normally distributed means that we don't expect extreme deviation from the mean random effect (which is zero)

## Confidence intervals

### Marginal prediction

```{r}
fish_marg_confint <- add_ci(
  fish_all_new, fish_lmm, 
  type="parametric",
  includeRanef=FALSE
)
fish_marg_confint
```

```{r}
ggplot(fish_sub) + 
  geom_ribbon(
    aes(x=Fish_Length, ymin=LCB0.025, ymax=UCB0.975), 
    data=fish_marg_confint,
    fill="gray", alpha=0.5
  ) + geom_line(
    aes(x=Fish_Length, y=pred), fish_marg_pred
  ) + geom_point(aes(
    x=Fish_Length, y=Trophic_Pos, 
    color=Lake, shape=Lake
  ))
```

### Conditional prediction

```{r}
fish_cond_confint <- add_ci(
  fish_new, fish_lmm, 
  type="parametric",
  includeRanef=TRUE
)
fish_cond_confint
```

```{r}
ggplot(fish_sub) + 
  geom_ribbon(
    aes(x=Fish_Length, ymin=LCB0.025, ymax=UCB0.975, fill=Lake, color=Lake), 
    data=fish_cond_confint,
    alpha=0.5
  ) + geom_line(
    aes(x=Fish_Length, y=pred, color=Lake), 
    fish_cond_pred
  ) + geom_point(aes(
    x=Fish_Length, y=Trophic_Pos, 
    color=Lake, shape=Lake
  )) + 
  geom_ribbon(
    aes(x=Fish_Length, ymin=LCB0.025, ymax=UCB0.975), 
    data=fish_marg_confint,
    fill="gray", alpha=0.5
  ) + geom_line(
    aes(x=Fish_Length, y=pred), fish_marg_pred
  ) + facet_wrap(vars(Lake))
```

## Prediction interval

### Marginal prediction

```{r}
fish_marg_predint <- add_pi(
  fish_all_new, fish_lmm, 
  type="parametric",
  includeRanef=FALSE
)
fish_marg_predint
```

```{r}
ggplot(fish_sub) + 
  geom_ribbon(
    aes(x=Fish_Length, ymin=LPB0.025, ymax=UPB0.975), 
    data=fish_marg_predint,
    fill="gray", alpha=0.5
  ) + geom_line(
    aes(x=Fish_Length, y=pred), fish_marg_pred
  ) + geom_point(aes(
    x=Fish_Length, y=Trophic_Pos, 
    color=Lake, shape=Lake
  ))
```

### Conditional prediction

```{r}
fish_cond_predint <- add_pi(
  fish_new, fish_lmm, 
  type="parametric",
  includeRanef=TRUE
)
fish_cond_predint
```

```{r}
ggplot(fish_sub) + 
  geom_ribbon(
    aes(x=Fish_Length, ymin=LPB0.025, ymax=UPB0.975, fill=Lake, color=Lake), 
    data=fish_cond_predint,
    alpha=0.5
  ) + geom_line(
    aes(x=Fish_Length, y=pred, color=Lake), 
    fish_cond_pred
  ) + geom_point(aes(
    x=Fish_Length, y=Trophic_Pos, 
    color=Lake, shape=Lake
  )) +
  geom_ribbon(
    aes(x=Fish_Length, ymin=LPB0.025, ymax=UPB0.975), 
    data=fish_marg_predint,
    fill="gray", alpha=0.5
  ) + geom_line(
    aes(x=Fish_Length, y=pred), fish_marg_pred
  ) + facet_wrap(vars(Lake))
```


